\texttt{thornado}-transport: a realizability-preserving IMEX scheme for two-moment neutrino transport

### 2018-09-25 Documents/M1/Astronum_2018/manuscript.tex
Updated the figures.
Updated introduction section and backup.

\begin{document}
\title{Constraint-Preserving IMEX Scheme for the Two-Moment Model of Fermion Transport}

@@

@@

\input{sections/intro}
\input{sections/TwoMoments}
\input{sections/SpacialDiscretization}
\input{sections/TimeIntegration}
\input{sections/Numerical}

>>>>

\begin{document}
\title{\texttt{thornado}-transport: a realizability-preserving IMEX scheme for two-moment neutrino transport}

@@

\author{Bronson Messer}
\address{Department of Physics and Astronomy, University of Tennessee Knoxville, TN 37996-1200}
\address{Scientific Computing and Theoretical Physics Groups, Oak Ridge National Laboratory, Oak Ridge, TN 37831 USA}
%\address{Physics Division, Oak Ridge National Laboratory, P.O. Box 2008, Oak Ridge, TN 37831-6354, USA}
\ead{bronson@ornl.gov}

@@

\input{sections/intro}
\input{sections/TwoMoments}
\input{sections/SpacialDiscretization}
\input{sections/TimeIntegration}
\input{sections/Numerical}
\input{sections/Conclusion}

### 2018-09-25 Documents/M1/Astronum_2018/sections/Numerical.tex
Updated introduction section and backup.

\section{Numerical Tests}
In this section we present the numerical results obtained with the PD-ARS schemes provided in this paper.
The tests in Section \ref{se: Accuracy Tests} are designed to compare the accuracy of the schemes in various regimes.
The test in Section \ref{se: Neutrino Stationary State Test} demonstrates the realizability-preserving of PD-ARSs scheme.
\subsection{Accuracy Tests}
\label{se: Accuracy Tests}
To compare the accuracy of the IMEX scheme, we applied our PDARSs and Pareschi \& Russo \cite{pareschiRusso_2005} (SSP2332) scheme to some smooth problems in streaming, absorption, ans scattering dominated regimes in one spatial dimension.
All the tests in this subsection were applied with third order accurate spatial discretization, the maximum entropy closure in the low occupancy limit and time step $\dt = 0.1 \times \dx $.
In the streaming test, the second-order and third-order accurate explicit strong stability-preserving Runge-Kutta methods\cite{gottlieb_etal_2001} , (SSPRK2 and SSPRK3, respectively) are also applied to serve as an reference.
More information and the definition of the absolute error and the relative error can be found in \cite{chu_2018}.

@@

\subsection{Neutrino Stationary State Test}
Next we consider a two dimension test with

\label{se: Neutrino Stationary State Test}
\begin{figure}[h]
  \centering
  \begin{tabular}{cc}
    \includegraphics[width=0.45\textwidth]{figures/NStatinaryS_EOS}
    \includegraphics[width=0.45\textwidth]{figures/NSS_Opacities}
  \end{tabular}
   \caption{Equation of state plot (left) and neutrino capture opacity (EC) and neutrino elastic scattering opacity (ES) in space (right) for the neutrino stationary state test.}
   \label{fig:NeutrinoStationaryTestEOS}
\end{figure}

\begin{figure}[h]
  \centering
    \includegraphics[width=\textwidth]{figures/NSS_1_1}\\
    \includegraphics[width=\textwidth]{figures/NSS_3_1} \\
    \includegraphics[width=\textwidth]{figures/NSS_5_1} \\
   \caption{Readability plots (left column), the number density $\cJ$ versus radius plots (center column) and the flux factor $|\bcH|/\cJ$ versus radius plots (right column) at $t = 0.01$ms ,$0.35$ms and $5.0$ms for the neutrino stationary state test.}
      \label{fig:NeutrinoStationaryTestEvolve}
\end{figure}

>>>>

\section{Numerical Tests}\label{se:NumericalTests}

In this section we present the numerical results obtained with the PD-ARS schemes provided in this paper.
The tests in Section \ref{se: Accuracy Tests} are designed to compare the accuracy of the schemes in various regimes.
The test in Section \ref{se: Neutrino Stationary State Test} demonstrates the realizability-preserving of PD-ARSs scheme.
All the tests in this subsection were applied with third-order accurate spatial discretization (polynomials of degree $k=2$) and time step $\dt = 0.1 \times \dx $.
\subsection{Accuracy Tests}
\label{se: Accuracy Tests}
To compare the accuracy of the IMEX scheme, we applied our PDARSs and Pareschi \& Russo \cite{pareschiRusso_2005} (SSP2332) scheme to some smooth problems in streaming, absorption, ans scattering dominated regimes in one spatial dimension.
All the tests in this subsection were applied with the maximum entropy closure in the low occupancy limit.
In the streaming test, the second-order and third-order accurate explicit strong stability-preserving Runge-Kutta methods\cite{gottlieb_etal_2001} , (SSPRK2 and SSPRK3, respectively) are also applied to serve as an reference.
More information and the definition of the absolute error and the relative error can be found in \cite{chu_2018}.

@@

\subsection{Neutrino Stationary State Test} \label{se: Neutrino Stationary State Test}
Next we consider a more realistic-like test: two-dimensional neutrino transport with emission, absorption and elastic scattering under a stationary realistic-like background.
This test is designed to test the realizability-preserving properties of the PD-ARSs schemes.
Figure~\ref{fig:NeutrinoStationaryTestEOS} plots the thermal state and the opacities we used.
This test is computed on a two-dimensional domain $D=\{\vect{x}\in\bbR^{2}:x^{1}\in[0,200], x^{2}\in[0,200]\}$ with a grid of 128 elements on each direction.
We take 10 energy groups vary from 0~MeV to 300~MeV and initial the neutrino number density $\cJ$ and flux density $\bcH$ both zero(arbitrary small).
For realizability-preserving test, CB closure with PD-ARSs schemes, SSP2332, scheme proposed by Cavaglieri \& Bewley \cite{cavaglieriBewley2015} and scheme proposed by McClarren et al. \cite{mcclarren_etal_2008} are all applied.
Only PD-ARSs schemes produces realizability-preserving result and evolve the model to a stable state.
The realizability-preserving plot and the evolve process of PD-ARSs are given in Figure~\ref{fig:NeutrinoStationaryTestEvolve}.

\begin{figure}[h]
  \centering
  \begin{tabular}{cc}
    \includegraphics[width=0.45\textwidth]{figures/NStatinaryS_EOS}
    \includegraphics[width=0.45\textwidth]{figures/NSS_Opacities}
  \end{tabular}
   \caption{Thermal state plot (left) and neutrino absorptivity ($\chi$) and neutrino elastic scattering opacity ($\sigma$) in space (right) for the neutrino stationary state test.}
   \label{fig:NeutrinoStationaryTestEOS}
\end{figure}

\begin{figure}[h]
  \centering
    \includegraphics[width=\textwidth]{figures/NSS_1_1}\\
    \includegraphics[width=\textwidth]{figures/NSS_3_1} \\
    \includegraphics[width=\textwidth]{figures/NSS_5_1} \\
   \caption{Readability plots (left column), the number density $\cJ$ versus radius plots (center column) and the flux factor $|\bcH|/\cJ$ versus radius plots (right column) at t = 0.01~ms ,0.35~ms and 5.0~ms for the neutrino stationary state test. For the realizability plots, the light blue area demonstrate the realizability domain and the black lines define its boundary. Each $\bcM=(\cJ,\bcH)^{T}$ state is marked by a red dot. The results of PD-ARS with SSPRK2 and PD-ARS with SSPRK3 are indistinguishable in these plot.}
      \label{fig:NeutrinoStationaryTestEvolve}
\end{figure}

### 2018-09-25 Documents/M1/Astronum_2018/sections/SpacialDiscretization.tex
Backup Astronum2018.
Updated introduction section and backup.

\section{Spacial Discretization}
Divide the spatial domain $D$ into $N$ cells and denote the cell as $I_{i}$ with $i = 1,\ldots,N$, so that
\begin{equation*}
D = \cup_{i = 1}^{N} I_{i}.
\end{equation*}
Each of the cell has a volume/length of $\dx$.
Then we have cell average moment:
\begin{equation}
\bcM_{i} = \dfrac{1}{\dx} \int_{I_i}\bcM dx.
\end{equation}
Integrate Eq.~\eqref{eq:momentEquations} for each cell $I_{i}$ to have
\begin{equation}
\dfrac{d \bcM_{i}}{d t} = - \dfrac{1}{\dx} \left( \widehat{\bcF}(\bcM_{i},\bcM_{i+1}) -  \widehat{\bcF}(\bcM_{i-1},\bcM_{i})\right) + \f{1}{\tau}\,\cC(\bcM_{i}),
\end{equation}
where $\widehat{\bcF}(\vect{\cM}_{a},\vect{\cM}_{b})$ is numerical flux and $\f{1}{\tau}\,\cC(\bcM_{i})$ is the integrated collision term.
In this paper we use the simple Lax-Friedrichs flux that given by
\begin{equation}
  \widehat{\bcF}_{LF}(\vect{\cM}_{a},\vect{\cM}_{b})
  =\f{1}{2}\,\big(\,\bcF(\vect{\cM}_{a})+\bcF(\vect{\cM}_{b})-(\,\vect{\cM}_{b}-\vect{\cM}_{a}\,)\,\big).
\end{equation}
Therefore, by treating the flux term explicitly and the collision term implicitly, we have
\begin{align}
\bcM_{i}^{n+1} 
& = \bcM_{i}^{n} - \dfrac{\dt}{\dx} \left( \widehat{\bcF}(\bcM^{n}_{i},\bcM^{n}_{i+1}) -  \widehat{\bcF}(\bcM^{n}_{i-1},\bcM^{n}_{i})\right) + \f{\dt}{\tau}\,\cC(\bcM^{n+1}_{i}) \nonumber \\
& = (1-\eta)\bcM_{i}^{n} + \f{1}{2}\eta\left( \bcM^{n}_{i+1}-\bcF(\bcM^{n}_{i+1})\right)  + \f{1}{2}\eta\left( \bcM^{n}_{i-1}+\bcF(\bcM^{n}_{i-1})\right) + \f{\dt}{\tau}\,\cC(\bcM^{n+1}_{i})
\end{align}
where $\eta := \dfrac{\dt}{\dx}$. The first three terms can form a 3-term convex combination by requiring $\eta \in [0,1]$.
It can be shown that $\bcM_{i}^{n+1}$ remains realizable given $\bcM_{i}^{n}$ is realizable, $\eta \in [0,1]$ and $\cC(\bcM)$ is positivity-preserving. The proof can be found in \cite{Chu_2018}.

>>>>

\section{Spacial Discretization}\label{se:SpacialDiscretization}

Divide the spatial domain $D$ into $N$ cells and denote the cell as $I_{i}$ with $i = 1,\ldots,N$, so that
\begin{equation*}
D = \cup_{i = 1}^{N} I_{i}.
\end{equation*}
Each of the cell has a volume/length of $\dx$.
Then we have cell average moment:
\begin{equation}
\bcM_{i} = \dfrac{1}{\dx} \int_{I_i}\bcM dx.
\end{equation}
Integrate Eq.~\eqref{eq:momentEquations} for each cell $I_{i}$ to have
\begin{equation}
\dfrac{d \bcM_{i}}{d t} = - \dfrac{1}{\dx} \left( \widehat{\bcF}(\bcM_{i},\bcM_{i+1}) -  \widehat{\bcF}(\bcM_{i-1},\bcM_{i})\right) + \f{1}{\tau}\,\cC(\bcM_{i}),
\end{equation}
where $\widehat{\bcF}(\vect{\cM}_{a},\vect{\cM}_{b})$ is numerical flux and $\f{1}{\tau}\,\cC(\bcM_{i})$ is the integrated collision term.
In this paper we use the simple Lax-Friedrichs flux that given by
\begin{equation}
  \widehat{\bcF}_{LF}(\vect{\cM}_{a},\vect{\cM}_{b})
  =\f{1}{2}\,\big(\,\bcF(\vect{\cM}_{a})+\bcF(\vect{\cM}_{b})-(\,\vect{\cM}_{b}-\vect{\cM}_{a}\,)\,\big).
\end{equation}
Therefore, by treating the flux term explicitly and the collision term implicitly, we have
\begin{align}
\bcM_{i}^{n+1} 
& = \bcM_{i}^{n} - \dfrac{\dt}{\dx} \left( \widehat{\bcF}(\bcM^{n}_{i},\bcM^{n}_{i+1}) -  \widehat{\bcF}(\bcM^{n}_{i-1},\bcM^{n}_{i})\right) + \f{\dt}{\tau}\,\cC(\bcM^{n+1}_{i}) \nonumber \\
& = (1-\eta)\bcM_{i}^{n} + \f{1}{2}\eta\left( \bcM^{n}_{i+1}-\bcF(\bcM^{n}_{i+1})\right)  + \f{1}{2}\eta\left( \bcM^{n}_{i-1}+\bcF(\bcM^{n}_{i-1})\right) + \f{\dt}{\tau}\,\cC(\bcM^{n+1}_{i})
\end{align}
where $\eta := \dfrac{\dt}{\dx}$. The first three terms can form a 3-term convex combination by requiring $\eta \in [0,1]$.
It can be shown that $\bcM_{i}^{n+1}$ remains realizable given $\bcM_{i}^{n}$ is realizable, $\eta \in [0,1]$ and $\cC(\bcM)$ is positivity-preserving. The proof can be found in \cite{c
chu_2018}.

### 2018-09-25 Documents/M1/Astronum_2018/sections/TimeIntegration.tex
Backup Astronum2018.
Updated introduction section and backup.

\section{Time Integration}
In this section we show how to use the property we introduced at the end of the previous section to construct a constraint-preserving implicit-explicit (IMEX) scheme.
We can have a more general form of the partial differential equation, Eq.~\eqref{eq:momentEquations},
\begin{equation}
  \dfrac{d \vect{u}}{d t} = \vect{\cT}(\vect{u}) + \f{1}{\tau}\,\vect{\cQ}(\vect{u}),
\end{equation}
where
\begin{equation}
\vect{u}(t) = \left( \bcM_{1}(t),\ldots,\bcM_{N}(t)\right) ^{T}
\end{equation}
is the collection of all moment cell average, $\vect{\cT}$ is the operator corresponding to the flux term and $\f{1}{\tau}\,\vect{\cQ}$ is a linear operators corresponding to the collision term.

@@

\subsection{Constraint-Preserving Implicit-Explicit Scheme}
The problem left for constructing a constraint-preserving IMEX is finding the conditions $(\tilde{a}_{ij})$, $(a_{ij})$, $\tilde{w}_{i}$ and $w_{i}$ need to satisfied and giving their possible values.
As Hu et al. did in their paper \cite{hu_etal_2018}, the stage values in Eq.~\eqref{imexStages} can be rewritten to
\begin{equation}
  \vect{u}^{(i)}
  =\sum_{j=0}^{i-1}c_{ij}\Big[\,\vect{u}^{(j)}+\hat{c}_{ij}\,\dt\,\vect{\cT}(\vect{u}^{(j)})\,\Big]
  +a_{ii}\,\dt\,\f{1}{\tau}\,\vect{\cQ}(\vect{u}^{(i)}),\quad i=1,\ldots,s,
  \label{eq:imexStagesRewrite}
\end{equation}
It can be proved \cite{Chu_2018} that the conditions we need for constraint-preserving IMEX is:
\begin{enumerate}
    \item Consistency of the implicit coefficients
    \begin{equation}
      \sum_{i=1}^{s}w_{i}=1.
    \end{equation}
    \item Second-order accuracy in the streaming limit; i.e., satisfies \begin{equation}
      \sum_{i=1}^{s}\tilde{w}_{i}=1
      \quad\text{and}\quad
      \sum_{i=1}^{s}\tilde{w}_{i}\,\tilde{c}_{i}=\f{1}{2},
      \label{eq:orderConditionsEx}
    \end{equation}
    \item Convex-invariant; i.e. satisfies 
    \begin{align}
      &a_{ii}>0, \quad c_{i0},\tilde{c}_{i0}\ge0, \quad \text{for} \quad i=2,\ldots,s, \nonumber \\
      &\text{and} \quad c_{ij},\tilde{c}_{ij}\ge0, \quad \text{for} \quad i=3,\ldots,s, \quad\text{and}\quad j=2,\ldots,i-1.  
    \end{align}
    with $\sum_{j=0}^{i-1}c_{ij}=1$, for $i=1,\ldots,s$, and $c_{\Sch}=\min_{\substack{i = 2,\ldots,s \\ 
                  j = 0,2,\ldots,i-1}}\,\f{1}{\hat{c}_{ij}}>0$.
    \item Well-behaved in the diffusion limit; i.e., satisfies \begin{equation}
      \vect{e}_{i}^{T}A^{-1}\tilde{A}\,\vect{e} = 1, \quad i=2,\ldots,s.
      \label{eq:diffusionCondition}
    \end{equation}
    \item Less than four stages ($s\le3$).
    \item Globally stiffly accurate (GSA): $a_{si}=w_{i}$ and $\tilde{a}_{si}=\tilde{w}_{i},\quad i=1,\ldots,s$.
\end{enumerate}  
We call the IMEX scheme satisfying the above conditions {PD-ARS}. (Definition 3 in \cite{Chu_2018})
We give two optimized PD-ARS scheme here: PD-ARS with SSPRK2 and  PD-ARS with SSPRK3. They have $2$-nd order and $3$-rd order accuracy in the streaming limit, respectively.
\subsubsection{PD-ARS with SSPRK2}
The explicit part of this IMEX scheme is SSPRK2.
\begin{equation}
  \begin{array}{c | c c c}
  	0 & 0   & 0 & 0 \\
  	1 & 1   & 0 & 0 \\
  	1 & 1/2 & 1/2 & 0 \\ \hline
  	  & 1/2 & 1/2 & 0
  \end{array}
  \qquad
  \begin{array}{c | c c c}
  	0 & 0 & 0            & 0            \\
  	1 & 0 & 1            & 0            \\
  	1 & 0 & 1/2-\epsilon & 1/2+\epsilon \\ \hline
  	  & 0 & 1/2-\epsilon & 1/2+\epsilon
  \end{array}
\end{equation}

>>>>

\section{Time Integration} \label{se:TimeIntegration}

In this section we show how to use the property we introduced at the end of the previous section to construct a constraint-preserving implicit-explicit (IMEX) scheme.
We can have a more general form of the partial differential equation, Eq.~\eqref{eq:momentEquations},
\begin{equation}
  \dfrac{d \vect{u}}{d t} = \vect{\cT}(\vect{u}) + \f{1}{\tau}\,\vect{\cQ}(\vect{u}),
\end{equation}
where
\begin{equation}
\vect{u}(t) = \left( \bcM_{1}(t),\ldots,\bcM_{N}(t)\right) ^{T}
\end{equation}
is the collection of all moment cell average, $\vect{\cT}$ is the operator corresponding to the flux term and $\f{1}{\tau}\,\vect{\cQ}$ is a linear operators corresponding to the collision term.

@@

\subsection{Constraint-Preserving Implicit-Explicit Scheme}
The problem left for constructing a constraint-preserving IMEX is finding the conditions $(\tilde{a}_{ij})$, $(a_{ij})$, $\tilde{w}_{i}$ and $w_{i}$ need to satisfied and giving their possible values.
As Hu et al. did in their paper \cite{hu_etal_2018}, the stage values in Eq.~\eqref{imexStages} can be rewritten to
\begin{equation}
  \vect{u}^{(i)}
  =\sum_{j=0}^{i-1}c_{ij}\Big[\,\vect{u}^{(j)}+\hat{c}_{ij}\,\dt\,\vect{\cT}(\vect{u}^{(j)})\,\Big]
  +a_{ii}\,\dt\,\f{1}{\tau}\,\vect{\cQ}(\vect{u}^{(i)}),\quad i=1,\ldots,s,
  \label{eq:imexStagesRewrite}
\end{equation}
It can be proved \cite{chu_2018} that the conditions we need for constraint-preserving IMEX is:
\begin{enumerate}
    \item Consistency of the implicit coefficients
    \begin{equation}
      \sum_{i=1}^{s}w_{i}=1.
    \end{equation}
    \item Second-order accuracy in the streaming limit; i.e., satisfies \begin{equation}
      \sum_{i=1}^{s}\tilde{w}_{i}=1
      \quad\text{and}\quad
      \sum_{i=1}^{s}\tilde{w}_{i}\,\tilde{c}_{i}=\f{1}{2},
      \label{eq:orderConditionsEx}
    \end{equation}
    \item Convex-invariant; i.e. satisfies 
    \begin{align}
      &a_{ii}>0, \quad c_{i0},\tilde{c}_{i0}\ge0, \quad \text{for} \quad i=2,\ldots,s, \nonumber \\
      &\text{and} \quad c_{ij},\tilde{c}_{ij}\ge0, \quad \text{for} \quad i=3,\ldots,s, \quad\text{and}\quad j=2,\ldots,i-1.  
    \end{align}
    with $\sum_{j=0}^{i-1}c_{ij}=1$, for $i=1,\ldots,s$, and $c_{\Sch}=\min_{\substack{i = 2,\ldots,s \\ 
                  j = 0,2,\ldots,i-1}}\,\f{1}{\hat{c}_{ij}}>0$.
    \item Well-behaved in the diffusion limit; i.e., satisfies \begin{equation}
      \vect{e}_{i}^{T}A^{-1}\tilde{A}\,\vect{e} = 1, \quad i=2,\ldots,s.
      \label{eq:diffusionCondition}
    \end{equation}
    \item Less than four stages ($s\le3$).
    \item Globally stiffly accurate (GSA): $a_{si}=w_{i}$ and $\tilde{a}_{si}=\tilde{w}_{i},\quad i=1,\ldots,s$.
\end{enumerate}  
We call the IMEX scheme satisfying the above conditions {PD-ARS}. (Definition 3 in \cite{Chu_2018})
We give two optimized PD-ARS scheme here: PD-ARS with SSPRK2 and  PD-ARS with SSPRK3. They have $2$-nd order and $3$-rd order accuracy in the streaming limit, respectively.
\subsubsection{PD-ARS with SSPRK2}
The explicit part of this IMEX scheme is SSPRK2.
\begin{equation}
  \begin{array}{c | c c c}
  	0 & 0   & 0 & 0 \\
  	1 & 1   & 0 & 0 \\
  	1 & 1/2 & 1/2 & 0 \\ \hline
  	  & 1/2 & 1/2 & 0
  \end{array}
  \qquad
  \begin{array}{c | c c c}
  	0 & 0 & 0            & 0            \\
  	1 & 0 & 1            & 0            \\
  	1 & 0 & 1/2-\epsilon & 1/2+\epsilon \\ \hline
  	  & 0 & 1/2-\epsilon & 1/2+\epsilon
  \end{array}
\end{equation}
